---
title: "{xplainfi} Benchmark Overview"
format: 
  revealjs:
    smaller: true
---

## xplainfi Benchmark Suite

-   Goals: 3 different scopes
      a)  **Sanity check** / regression tests for `xplainfi` implementations
      b)  **Runtime** comparison, performance analysis
      c)  Comparing **importance** results + **runtime** between implementations for paper   
          --> combination / subset of a + b
-   Different algorithm / problem combinations fit different scopes
- Priority: What's needed for paper

##   General components

- Standard batchtools + mlr3 setup: Problems & algorithms
-   **Problem instance**: 
    - Fixed: 
      - Simulated/real data/task  (instantiated resampling (holdout))
        - `Parameter`: n (p, correlation structure, ...)
        - Currently regression tasks only (sufficient?)
      - Measure (MSE for regr (acc for classif))
      - `Parameter`: Learner/model class for train/predict step
-   **Algorithm**: 
    -  FI method from specific package
    -  `Parameter`: Conditional sampler

## Algorithms to include

Focus on **model agnostic** methods and implementations in R (/python)

|            |            |        *Software*   |   |    |       |
|------------|------------|--------------|-------|-------|-----------------|
| *Method*   | `xplainfi` | `fippy` (py) | `iml` | `vip` | `sage` (py)     |
| **PFI**    | ✅         | ❌           | ✅    | ❓    | ---             |
| **CFI**    | ✅         | ✅           | ---   | ---   | ---             |
| **RFI**    | ❓         | ❓           | ---   | ---   | ---             |
| **LOCO**   | ✅         | ---          | ---   | ---   | ---             |
| **mSAGE**  | ✅         | ✅           | ---   | ---   | ✅ (KernelSAGE) |
| **cSAGE**  | ✅         | ✅           | ---   | ---   | ---             |


- Python implementations set up via `reticulate`
- ✅ include, ❌ exclude, ❓ unsure, --- not implemented


## Algorithm part: Model classes

| Class         | R           | Python  | Purpose                                         |
|---------------|-------------|---------|-------------------------------------------------|
| Featureless   | featureless | ---     | Runtime overhead in xplainfi only               |
| Linear        | lm/log_reg  | sklearn | "Basic", misses interactions                    |
| RF (untuned)  | ranger      | sklearn | Picks up interactions, correlations problematic |
| MLP (untuned) | nnet        | sklearn | People will ask for this anyway                 |


## Algorithm part: Samplers

| Type          | xplainfi | fippy   | Scope           |
|---------------|----------|---------|-----------------|
| **Gaussian**  | ✅  | ✅ | Compare xplainfi / fippy       |
| kNN           | ✅  | --      | Everything else |
| ctree         | ✅  | ---     | Everything else |
| ARF           | ✅  | ---     | Everything else |
| Normal. Flows | ---      | ✅ | Can't compare in R            |
| RF-based univariate cont./cat.  | ---   | ✅ |  Categorical features, used in Marginal SAGE  |
<!-- | Sequential    | --       | ✅ |  Combine multiple samplers   | -->

- ✅ implemented, --- not implemented



## Problems / DGPs for paper

|DGP             |Challenge               |Goal / scope           |
|:--------------|:------------------------|:---------------------|
| ✅ `ewald`        | Mixed effects            | Compare methods in general |
| ✅ `correlated`   | Spurious correlation     | Marginal vs. conditional |
| ✅ `interactions` | Interaction effects      | Dependency on model class |
| ✅ `mlbench.peak` | varying n, p           | **Runtime** only |
| ✅ `bike_sharing`   | somewhat large / wide   | Real data, well known   |
| ❓ `independent`  | Baseline (no challenges) | Sanity check? |
| ❓ `friedman1`   | *Nonlinear* DGP, 5 causal / noise | Model class dependency? |
| ❓ `confounded`   | Confounding              | Unsure |
| ❓ `mediated`     | Mediation effects        | Unsure |

Most tasks: p <= 5, linear DGP

<!-- ## Unknowns

- Categorical feature handling (bike_sharing)?
  - In R no encoding needed, in Python> one-hot encoding
  - -> FI not comparable, not sure
  - Worth the effort? 🫠
-  -->



## Runtime

- Mainly via `mlbench.peak` across different n, p
- `bike_sharing` large
- Caveat:
  - `xplainfi` faster with `learner$predict_newdata_fast`
  - -> Not available if learner is `GraphLearner`, in general
  - Overhead is understood, setup here is without graphs, so optimistic scenario
- Conditional samplers: large effect depending on data size
- Pretests with
  - learners, problems, samplers as shown
  - `n_repeats` in `{1, 10}`
  - n_samples in `{100, 500, 1000}` (+ bike_sharing)

```{r read-data}
res = readRDS("runtime-pretest.rds")
library(ggplot2)
library(dplyr)
```

## Runtime: median (max.) minutes {.scrollable}


```{r runtime-minutes}
res |>
	filter(problem %in% c("peak", "bike_sharing", "ewald")) |>
	# filter(n_repeats == max(n_repeats) | is.na(n_repeats)) |>
	mutate(
		# problem = glue::glue("{problem} ({n_samples}⨉{n_features})"),
		method = case_when(
			stringr::str_detect(algorithm, "^(ConditionalSAGE|CFI)$") ~ glue::glue(
				"{algorithm} ({sampler})",
				.na = ""
			),
			TRUE ~ algorithm
		),
		runtime = runtime / 60
	) |>
	group_by(method, problem, learner_type) |>
	summarize(
		q25 = quantile(runtime, probs = 0.25),
		median = median(runtime),
		q75 = quantile(runtime, 0.75),
		max = max(runtime)
	) |>
	mutate(across(where(is.numeric), \(x) round(x, 1))) |>
	mutate(
		fmt = as.character(glue::glue("{median} ({max})"))
	) |>
	arrange(desc(median)) |>
	tidyr::pivot_wider(
		id_cols = c("method", "learner_type"),
		values_from = "fmt",
		names_from = "problem"
	) |>
	# reactable::reactable(sortable = TRUE, filterable = TRUE)
	kableExtra::kbl() |>
	kableExtra::kable_styling()
```


## Runtime: `mlbench.peak` {.nostretch .scrollable}


```{r peak-runtime}
#| fig-width: 10
#| fig-height: 7
#| fig-align: center
res |>
	dplyr::filter(
		problem == "peak",
		learner_type == "rf"
		# is.na(sampler) | sampler == "arf",
	) |>
	dplyr::mutate(
		method = ifelse(!is.na(sampler), glue::glue("{algorithm} ({sampler})"), algorithm)
	) |>
	ggplot(aes(y = reorder(method, runtime), x = runtime / 60 / 60)) +
	facet_wrap(vars(n_samples, n_features), labeller = label_both) +
	geom_boxplot() +
	scale_x_log10(labels = scales::label_log()) +
	labs(
		# title = "mlbench.peak with varying n, p",
		subtitle = "RF, 1 and 10 repeats",
		x = "Runtime (log 10 hours)",
		y = "Algorithm"
	) +
	theme_bw(base_size = 13) +
	theme(
		plot.title.position = "plot",
		axis.text.y = element_text(size = 7.5),
		strip.text = element_text(size = rel(0.7))
	)
```

## Runtime: `bike_sharing` {.nostretch .scrollable}

```{r bikesharing-runtime}
#| fig-width: 10
#| fig-height: 6
#| fig-align: center
res |>
	dplyr::filter(problem == "bike_sharing") |>
	dplyr::mutate(
		method = ifelse(!is.na(sampler), glue::glue("{algorithm} ({sampler})"), algorithm)
	) |>
	ggplot(aes(y = reorder(method, runtime), x = runtime / 60 / 60)) +
	facet_wrap(vars(learner_type)) +
	geom_boxplot() +
	scale_x_log10(labels = scales::label_log()) +
	labs(
		# title = "Runtime per algorithm, learner",
		x = "Runtime (log10 hours)",
		y = NULL
	) +
	theme_bw(base_size = 13) +
	theme(plot.title.position = "plot")
```
